Characterizing the response of subarctic European vegetation to key environmental variables for land surface modeling of ozone pollution and risk assessment

### 2021-10-14 VegOzone.tex
Include Ane's suggestions

Ground-level ozone (\chem{O_3}) is a highly toxic pollutant known to cause damage to human health \citep{WHO2008} and to a variety of ecosystems around the World \citep{PT:Emberson2020}. Ozone is a cause of reduction of photosynthesis, visible foliar injury, early senescence and programmed cell death of plants \citep{PCE:Kangasjarvi2005}. Annual global yield losses of four major crops (wheat, rice, maize, and soybean) of $3-15\,\unit{\%}$ have been reported \citep{PJ:Ainsworth2017} and a suggested loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone.
The crop yield loss studies indicate a threat to food security in rapidly developing countries, e.g., in East and South-East Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018}, where ozone concentrations related to enhanced pollutant emissions from the transport, residential, power generation, and industrial sectors have increased in the past few decades. Even though long-term (pre-1950) observations of ground-level ozone concentrations are scarce; evidence from trend analysis of those few sites suggests that since the industrial revolution, background concentrations in the northern hemisphere have at least doubled and continued to increase \citep{IPCC2013}. Recent trend analyses for Europe indicate that a maximum was reached in 2007 \citep{AE:Derwent2018}, after which tropospheric background ozone started to level off or even decline \citep{ESA:Cooper2014, ACP:Wespes2018,ESA:Gaudel2018}. This can be attributed to a successful implementation of air quality regulations coupled with economic restructuring, reducing the number of episodes of peak concentrations especially in summertime \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}. However, changes in environmental conditions associated with climate change, such as an increase in the frequency of heatwaves, could negate the effectiveness of emission reductions in relation to ozone impacts and air quality \citep{NCC:Lin2020}.

@@

In the annual cycle in Boreal climates, a distinct maximum of ground-level ozone concentrations typically occurs in spring followed by lower concentrations throughout the summer. Several conditions unique to the Arctic are stimulating the formation of this spring peak. Major pathways of ozone and its precursors' removal from the boundary layer include dry deposition on vegetated surfaces and photochemistry \citep{RG:Clifton2020}. As dry deposition to bare ground or snow and ice-covered surfaces was found to be low \citep{ACP:Helmig2007}, both pathways of removal are substantially suppressed during polar night \citep{AE:Monks2000}. This leads to a build-up of ozone and its precursors during winter. Come spring precursors are photochemically activated again, while snow cover typically continues to prevail and suppress ozone dry deposition.
On the other hand, very high ozone concentrations occurring over shorter periods can often be attributed to deep tropopause folding events over the Arctic during winter. These events facilitate an intrusion of stratospheric air masses which are enriched in ozone \citep{JGR:Skerlak2015}.
The spring peak is more pronounced in northern Fennoscandia than in more southern locations but shows a high interannual variability \citep{AB:Klingberg2009, BER:Klingberg2019}. A cause for this is the complex interplay between atmospheric chemistry and dynamics \citep{AE:Laurila1996,BER:Hatakka2003} as well as anthropogenic and biogenic activity \citep{AE:Rummukainen1996,AE:Simpson2002,QJRMS:Galbally2007,NGS:Schnell2009}. \citet{ACP:Andersson2017} showed in a modeling study focused on Fennoscandia that the variability of ozone concentrations in winter can be attributed mainly to changes in atmospheric background (transport) of ozone, while summertime abundance is mostly affected by emissions of precursors in the rest of Europe. Meteorology plays an important role throughout the year.

@@

To estimate the flux-based metric, also referred to as the Phytotoxic Ozone Dose over a threshold y ($\mathrm{POD_y}$), an estimate of the stomatal \chem{O_3} flux ($\Phi_\mathrm{sto}$) is calculated based on the assumption that the concentration of \chem{O_3} at the top of the canopy represents a reasonable estimate of the concentration at the upper surface of the laminar layer for a sunlit upper canopy leaf and environmental factors affecting stomatal conductance. $\mathrm{POD_y}$ is then calculated according to:
%
\begin{equation}
  \mathrm{POD_y} = \int{(\Phi_\mathrm{sto}-y)\cdot \mathrm{dt}},
  \label{eq:pod}
\end{equation}
%
with the hourly averaged stomatal ozone flux $\Phi_\mathrm{sto}$ (see Eq.~(\ref{eq:flux_stomata})) and a stomatal ozone flux threshold $\mathrm{y}$ both given in units of \unit{nmol\,\chem{O_3}\,m^{-2}\,\mathrm{PLA}\,s^{-1}}. The flux threshold $\mathrm{y}$ represents the detoxification potential of the plant and is typically only exceeded during daylight hours (i.e. when global radiation is above $50\,\unit{W\,m^{-2}}$). $\mathrm{POD_y}$ is given in \unit{mmol\,m^{-2}} integrated over the growing season.

The $\mathrm{POD_y}$ metric is currently used in risk assessments under the United Nations Economic Commission for Europe (UNECE) Long Range Transboundary Air Pollution (LRTAP) Convention to identify those locations across Europe where vegetation is at risk from ozone. This convention aims to develop an effects-based emission reduction policy that can target those ozone precursor emissions that are most influential in causing damage. To achieve this the concept of critical levels (CLs) has been developed and applied \citep{Maas2016}. Exceedance of the CL is used to identify those areas across Europe that would benefit from targeted Europe-wide emission reductions. Methods to estimate $\mathrm{POD_y}$ and CLs are defined by the UNECE LRTAP Convention and described in a Mapping Manual \citep{ICP:MappingManual2017}. The CL is calculated by: 
%
\begin{equation}
  \mathrm{CL_{exeed}} = \mathrm{POD_y} - \mathrm{CL}.
\end{equation}
%

An application of the Mapping Manual method was made by \citet{GCB:Mills2011} where exceedance of the CL was related to observations of ozone damage to a clover bio-monitor. This study demonstrated the improved ability of flux-based metrics (in comparison to exposure-based metrics) to identify the geographical distribution of the risk of ozone damage and also showed that the ozone damage could extend into more northerly regions of Europe. For our study, we conducted a similar bio-monitoring study at the Norwegian Institute of Bioeconomy Research (NIBIO) Environment Centre Svanhovd to assess whether local ozone concentrations were capable of inducing damage to vegetation growing under northern Fennoscandia conditions. In 2018, we indeed observed ozone damage on clover and tobacco. In contrast, no such damage was found on the clovers in 2019. This indicates that environmental factors in 2018 differed from 2019.

According to a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record -- favorable conditions for ground-level ozone formation.
An unusually weak and northward shifted jet stream allowed for a persistent high-pressure system above northern Europe, including Fennoscandia which blocked the low-pressure systems for several consecutive months. In the period May -- July, southern Norway experienced temperatures $4\,\unit{^\circ C}$ above normal and saw only about $60\,\unit{\%}$ of normal precipitation.Northern Norway as a whole had close to normal precipitation, but with local variations.
Thermal stress on vegetation was exceptional not only in large parts of Fennoscandia but also in much of Europe, where the influence of the high-pressure system extended even over five months (April/May, July--September). These conditions gave rise to massive forest fires in different parts of Europe and thus an increase in ozone precursors.
A total of 2079 forest fires were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep{DSB2019}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer), and an estimated total of $25000\,\unit{hectare}$ burned down in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) \citep{SOU2019}. Boreal wildfires are found to emit ozone precursors \chem{CO} ($\chem{[CO]}/\chem{[CO_2]} \propto 6-13\,\unit{\%}$) and VOCs ($\chem{[VOCs]}/\chem{[CO_2]} \propto 0.5-1.5\,\unit{\%}$) \citep{AE:Cofer1990}. Coincident peak \chem{[O_3]} is found in ozone monitoring data (Fig.~\ref{fig:data_svanvik_2018_2019}) in July. It is very likely that elevated ozone, like during the 2003 drought period \citep{JGR:Solberg2018}, was promoted by a combination of various factors such as wildfires, reduced cloud cover (increased solar radiation), reduced dry deposition and turbulent mixing due to the stagnant weather conditions, and increased BVOC emissions. As such events may become more frequent in the future, conditions at Svanhovd for the growing seasons 2018/19 shall serve as a reference for present and probable future conditions in northern Fennoscandia.\\

The described methods for risk assessment rely on accurate representation of plant physiological response to key environmental variables and are most often parameterized using data for less extreme climates, i.e., temperate or continental climates; and to a lesser extent, Boreal and Mediterranean climates. Parameterization of these models for more extreme bio-geographical regions such as subarctic climates has not been performed and relies on assuming that less extreme Boreal parameterization will be applicable. It is important to understand the implications of using these more generic parameterizations for more extreme climate regions. In our study, we explore how risk might be over- or underestimated using generic vs bespoke subarctic parameterization. We also investigate the effects of the interannual variability of essential climate variables (sensu World Meteorological Organization) on future risk of vegetation damage.
This allows an understanding of how the threat from ozone may change in the future and which are the key aspects of plant physiology that might determine the potential risk. Hence, we determine which parameters require special attention to define accurately for pollution impact assessment and more generally for describing gas exchange response to a changing climate that will influence biochemical cycling. This might also be one key to solve long-standing issues of earth system models (ESM) which often lack PFTs specifically representing subarctic conditions in their land surface models \citep{GMD:Poulter2015,JAMES:Lawrence2019}.

The observation site at NIBIO Environment Centre Svanhovd and available observational data are presented in Section~\ref{sec:data}. From these data, we derive multi-annual means of essential climate variables (temperature, precipitation, global irradiance, and ozone) and evaluate the deviation of 2018/19 from this norm (Section~\ref{sec:stats}). In Section~\ref{sec:do3se}, we use the $\mathrm{DO_3SE}$ model to estimate the ozone uptake by natural and semi-natural vegetation to investigate the main environmental drivers for projected ozone damage risk in 2018. To this end, we derive bespoke parameterizations for subarctic species to assess over- and underestimations arising from the use of generic parameterizations established from less climatically extreme conditions. We will summarize our results in Section~\ref{sec:conc} and discuss how this research can support improvements in future assessments of pollution risk and more generally biogeochemical cycling.

@@

In our assessment of primary drivers of ozone damage risk on vegetation in the subarctic, we focus on conditions at the NIBIO operated Environment Centre Svanhovd (Sør-Varanger, NOR). The NIBIO Environment Centre Svanhovd focuses on the research of the special environment of the region, but also comprises long-term observations of various agrometeorological variables, i.e. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths. We investigate the 2018/19 growing season in a climatological context. To this end, we collected atmospheric and ozone monitoring data taken at Svanhovd over the past $35\,\unit{years}$. Svanhovd synonymously refers to the atmospheric monitoring site Svanvik/Pasvik throughout the paper.\\

All measured essential climate variables including ozone for the 2018/19 growing season were obtained through \href{luftkvalitet.no}{luftkvalitet.no} operated by \citet{NILU_AIRQ} (last accessed May 2020). The ozone monitoring data have been also added to the EBAS database operated by \citet{NILU_EBAS} and are thus openly accessible. Long-term ozone observation data (1986--1996) were obtained from EBAS. Note that data from before the 1990s do not follow the high-quality standard procedures implemented nowadays and have to be treated with care \citep{NILU2003}. Especially, ozone monitors did not undergo regular re-calibration. This probably led to drifts in the observed data and may impose a false trend which shall not be of concern for us.
The long-term data sets of agrometeorological variables including temperature and precipitation are available from September 1992 to the present day \citep[LandbruksMeteorologiske Tjeneste][note the station name here is Pasvik]{LMT_NIBIO}.

To relate local ozone concentration \chem{[O_3]} with visible damage on ozone sensitive plants, both an ozone bio-monitor (referred to as ozone garden) and a conventional ozone monitor using the existing atmospheric monitoring infrastructure had been installed during the 2018/19 growing season at the NIBIO centre. We report \chem{[O_3]} as volume mixing ratios (VMR) in \unit{ppb} for easier comparison with exposure-based risk assessment. The locations of the atmospheric monitoring site and the ozone garden are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a. An ozone garden consists of selected plant species that are sensitive to ozone and are likely to display visible injuries. Cultivated species were, e.g. clover, tobacco, and potatoes. As shown in Fig.~\ref{fig:svanhovd_research_station}b, the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we qualitatively observed visible ozone damage on semi-natural vegetation (clover) and crops (tobacco). In contrast, no such visible damage was found on the clovers in 2019, although it was observed on the sensitive tobacco cultivar. This indicates that the vegetation was more affected by ozone in 2018 than in 2019.

@@

\section{Statistical analysis}
\label{sec:stats}
In this section, we assess the climate conditions at Svanhovd. To this end, we compute a multi-annual mean (referred to as climatology) for each environmental key variable (temperature, precipitation, global irradiance) and ground-level ozone. We evaluate the statistical significance of divergences from the norm in these variables (referred to as anomalies) in 2018/19. Concerning ozone, we performed a bias correction accounting for an increase in tropospheric background \chem{[O_3]} and cross-calibrated the climatology with long-term observations at other sites in northern Fennoscandia as described in \citet{ACPD:Falk2021}.

@@

The location of the NIBIO Environment Centre Svanhovd ($69.45\,\unit{^\circ N}$, $30.03\,\unit{^\circ E}$, $30\,\unit{m}$) suggests a subarctic climate. The 1992--2012 climatologies of temperature and precipitation are shown in a combined climate diagram (Fig.~\ref{fig:climatediagram}). Monthly averaged $2\,\unit{m}$ temperatures (red line) are displayed with a standard error of the mean (SE) as error band. We chose SE in this case due to the large interannual variability in temperature reflected in the standard deviation. Averaged monthly accumulated precipitation (blue bars) is shown with standard deviation as error bars. Months with average temperatures below freezing are denoted with a star. As can be seen, temperatures stay below freezing for 5 consecutive months, while only 2 months breach $10\,\unit{^\circ C}$ regularly (July, August), satisfying the conditions for K\"{o}ppen's climate classification of a regular subarctic climate (Dfc) \citep[e.g.][]{SD:Beck2018}. The highest monthly average temperature is $(13.1\pm 1.1)\,\unit{^\circ C}$ in July and the lowest $(-12.8\pm 2.0)\,\unit{^\circ C}$ in February. The coldest measured temperature was $-45.2\,\unit{^\circ C}$ in January 27, 1999, while the highest temperature ($29.4\,\unit{^\circ C}$) occurred in July 16 the same year.

@@

The average accumulated monthly precipitation ($\sum_d\mathrm{Precip}$) indicates that winter and spring (November--April, except for March) are relatively dry ($\sum_d\mathrm{Precip} < 20\,\unit{mm}$). The driest month is January with $\sum_d\mathrm{Precip} = (16.7\pm 3.0)\,\unit{mm}$. The most precipitation occurs in the summer months, with a $\sum_d\mathrm{Precip} = (58.5\pm 9.2)\,\unit{mm}$ in August. The average annual accumulated precipitation ($\sum_m\mathrm{Precip}$) given with standard error of mean is $(383\pm 86)\,\unit{mm}$. Precipitation shows a considerable interannual variability.

The climatology for global irradiance ($Q_0$, not shown) approaches zero in December/January (polar night) and reaches its maximum in June/July (midnight sun conditions). The highest observed averaged daily maximum ($\left<Q_0^\mathrm{max}\right>$) amounts to $800\,\unit{W\,m^{-2}}$, while the maximum of the averaged daily mean ($\left<Q_0^\mathrm{mean}\right>$) is $450\,\unit{W\,m^{-2}}$ (regularly reached in May--July). The maximum of the averaged daily minimum ($\left<Q_0^\mathrm{min}\right>$) lies at about $200\,\unit{W\,m^{-2}}$.

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{ozone_climatology_svanvik_obs_spline}
  %\includegraphics[width=8.3cm]{fig05}
  \caption{Svanhovd ozone climatology. The density distribution is shown together with a multi-annual mean of daily \chem{[O_3]} (climatology). A spline has been fitted through daily mean \chem{[O_3]}. The climatology of daily mean ozone is shown together with $1\,\sigma$ standard deviation and standard error of the mean.}
  \label{fig:ozone_climatology_fenoscandic_obs_spline}
\end{figure}

@@

We deduced the dominant species on-site at the NIBIO Environment Centre Svanhovd and from Fig.~\ref{fig:svanhovd_research_station} and found perennial grassland, birch (generalized as deciduous trees), and Scots pine (generalized as coniferous trees). Generic parameters for these PFTs are derived from \citet{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017} and will be referred to as mapping manual (MM) parameterizations in the following. Our parameterization of coniferous trees is based on the MM's Boreal Norway spruce, deciduous trees on the MM's silver birch, and perennial grassland on MM's perennial grassland for central Europe. For a comprehensive list of generic MM model parameters, consult Supplement~Table~S1.

@@

\begin{acknowledgements}
  Bj{\o}rg Rognerud for processing SeNorge.no data with respect for begin of growing season.
  Tore Flatlandsmo Berglen for hourly pressure data from Svanhovd.
  This work was supported by the Research Council of Norway (NRC) grand No. 268073.
\end{acknowledgements}

>>>>

Ground-level ozone (\chem{O_3}) is a highly toxic pollutant known to cause damage to human health \citep{WHO2008} and to a variety of ecosystems around the World \citep{PT:Emberson2020}. Ozone causes reduced photosynthesis, visible foliar injury, early senescence and programmed cell death of plants \citep{PCE:Kangasjarvi2005}. Annual global yield losses of four major crops (wheat, rice, maize, and soybean) of $3-15\,\unit{\%}$ have been reported \citep{PJ:Ainsworth2017} and a suggested loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone.
The crop yield loss studies indicate a threat to food security in rapidly developing countries, e.g., in East and South-East Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018}, where ozone concentrations related to enhanced pollutant emissions from the transport, residential, power generation, and industrial sectors have increased in the past few decades. Even though long-term (pre-1950) observations of ground-level ozone concentrations are scarce; evidence from trend analysis of those few sites suggests that since the industrial revolution, background concentrations in the northern hemisphere have at least doubled and continued to increase \citep{IPCC2013}. Recent trend analyses for Europe indicate that a maximum was reached in 2007 \citep{AE:Derwent2018}, after which tropospheric background ozone started to level off or even decline \citep{ESA:Cooper2014, ACP:Wespes2018,ESA:Gaudel2018}. This can be attributed to a successful implementation of air quality regulations coupled with economic restructuring, reducing the number of episodes of peak concentrations especially in summertime \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}. However, changes in environmental conditions associated with climate change, such as an increase in the frequency of heatwaves, could negate the effectiveness of emission reductions in relation to ozone impacts and air quality \citep{NCC:Lin2020}. Tropospheric background ozone concentrations derived from satellite instruments range between $50-65\,\unit{ppb}$ over Fennoscandia but strongly depend on season \citep{ESA:Cooper2014}.

@@

In the annual cycle in Boreal climates, a distinct maximum of ground-level ozone concentrations typically occurs in spring followed by lower concentrations throughout the summer. Several conditions unique to the Arctic are stimulating the formation of this spring peak. Major pathways of ozone and its precursors' removal from the boundary layer include dry deposition on vegetated surfaces and photochemistry \citep{RG:Clifton2020}. Dry deposition to bare ground or snow and ice-covered surfaces was found to be low \citep{ACP:Helmig2007}. This causes a  substantial suppression of ozone removal \citep{AE:Monks2000} and leads to a build-up of ozone and its precursors during winter. Come spring precursors are photochemically activated again, while snow cover typically continues to prevail and suppress ozone dry deposition.
On the other hand, very high ozone concentrations occurring over shorter periods can often be attributed to deep tropopause folding events over the Arctic during winter. These events facilitate an intrusion of stratospheric air masses which are enriched in ozone \citep{JGR:Skerlak2015}.
The spring peak is more pronounced in northern Fennoscandia than in more southern locations but shows a high interannual variability \citep{AB:Klingberg2009, BER:Klingberg2019}. \citet{ACP:Andersson2017} showed in a modeling study focused on Fennoscandia that the variability of ozone concentrations in winter can be attributed mainly to changes in atmospheric background (transport) of ozone, while summertime abundance is mostly affected by emissions of precursors in the rest of Europe.

@@

To estimate the flux-based metric, also referred to as the Phytotoxic Ozone Dose over a threshold y ($\mathrm{POD_y}$), an estimate of the stomatal \chem{O_3} flux ($\Phi_\mathrm{sto}$) is calculated based on the assumption that the concentration of \chem{O_3} at the top of the canopy represents a reasonable estimate of the concentration at the upper surface of the laminar layer for a sunlit upper canopy leaf and environmental factors affecting stomatal conductance. $\mathrm{POD_y}$ is then calculated according to:
%
\begin{equation}
  \mathrm{POD_y} = \int{(\Phi_\mathrm{sto}-y)\cdot \mathrm{dt}},
  \label{eq:pod}
\end{equation}
%
with the hourly averaged stomatal ozone flux $\Phi_\mathrm{sto}$ (see Appendix Eq.~(\ref{eq:flux_stomata}) and a stomatal ozone flux threshold $\mathrm{y}$ both given in units of \unit{nmol\,\chem{O_3}\,m^{-2}\,\mathrm{PLA}\,s^{-1}}. The flux threshold $\mathrm{y}$ represents the detoxification potential of the plant and is typically only exceeded during daylight hours (i.e. when global radiation is above $50\,\unit{W\,m^{-2}}$). $\mathrm{POD_y}$ is given in \unit{mmol\,m^{-2}} integrated over the growing season.

The $\mathrm{POD_y}$ metric is currently used in risk assessments under the United Nations Economic Commission for Europe (UNECE) Long Range Transboundary Air Pollution (LRTAP) Convention to identify those locations across Europe where vegetation is at risk from ozone. This convention aims to develop an effects-based emission reduction policy that can target those ozone precursor emissions that are most influential in causing damage. To achieve this, the concept of critical levels (CLs) has been developed and applied \citep{Maas2016}. Exceedance of the CL is used to identify those areas across Europe that would benefit from targeted Europe-wide emission reductions. Methods to estimate $\mathrm{POD_y}$ and CLs are defined by the UNECE LRTAP Convention and described in a Mapping Manual \citep{ICP:MappingManual2017}. The CL is calculated by: 
%
\begin{equation}
  \mathrm{CL_{exeed}} = \mathrm{POD_y} - \mathrm{CL}.
\end{equation}
%

An application of the Mapping Manual method was made by \citet{GCB:Mills2011} where exceedance of the CL was related to observations of ozone damage to a clover bio-monitor. This study demonstrated the improved ability of flux-based metrics (in comparison to exposure-based metrics) to identify the geographical distribution of the risk of ozone damage and also showed that the ozone damage could extend into more northerly regions of Europe. For our study, we conducted a similar bio-monitoring study at Svanhovd, a site with agrometeorological and air pollution monitoring, to assess whether local ozone concentrations were capable of inducing damage to vegetation growing under northern Fennoscandia conditions. In 2018, we indeed observed ozone damage on clover and tobacco. In contrast, no such damage was found on the clovers in 2019. This indicates that environmental factors in 2018 differed from 2019.

According to a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record -- favorable conditions for ground-level ozone formation.
An unusually weak and northward shifted jet stream allowed for a persistent high-pressure system above northern Europe, including Fennoscandia which blocked the low-pressure systems for several consecutive months. In the period May -- July, southern Norway experienced temperatures $4\,\unit{^\circ C}$ above normal and saw only about $60\,\unit{\%}$ of normal precipitation. Northern Norway as a whole had close to normal precipitation, one of the causes of the high rate of forest fires in that area.
Thermal stress on vegetation was exceptional not only in large parts of Fennoscandia but also in much of Europe, where the influence of the high-pressure system extended even over five months (April/May, July--September). These conditions gave rise to massive forest fires in different parts of Europe and thus an increase in ozone precursors.
A total of 2079 forest fires were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep{DSB2019}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer), and an estimated total of $25000\,\unit{hectare}$ burned down in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) \citep{SOU2019}. Boreal wildfires are found to emit ozone precursors \chem{CO} ($\chem{[CO]}/\chem{[CO_2]} \propto 6-13\,\unit{\%}$) and VOCs ($\chem{[VOCs]}/\chem{[CO_2]} \propto 0.5-1.5\,\unit{\%}$) \citep{AE:Cofer1990}. Coincident peak \chem{[O_3]} is found in ozone monitoring data (Fig.~\ref{fig:data_svanvik_2018_2019}) in July. It is very likely that elevated ozone, like during the 2003 drought period \citep{JGR:Solberg2018}, was promoted by a combination of various factors such as wildfires, reduced cloud cover (increased solar radiation), reduced dry deposition and turbulent mixing due to the stagnant weather conditions, and increased BVOC emissions. As such events may become more frequent in the future, conditions at Svanhovd for the growing seasons 2018 shall serve as a reference for probable future conditions in northern Fennoscandia. Conditions in 2019 represent present climate.\\

The described methods for risk assessment rely on accurate representation of plant physiological responses to key environmental variables and are most often parameterized using data for less extreme climates, i.e., temperate or continental climates; and to a lesser extent, Boreal and Mediterranean climates. Parameterization of these models for more extreme bio-geographical regions such as subarctic climates has not been performed and relies on assuming that less extreme Boreal parameterization will be applicable. It is important to understand the implications of using these more generic parameterizations for more extreme climate regions. In our study, we explore how risk might be over- or underestimated using generic vs bespoke subarctic parameterization. We also investigate the effects of the interannual variability of essential climate variables (sensu World Meteorological Organization) on future risk of vegetation damage.
This allows an understanding of how the threat from ozone may change in the future and which are the key aspects of plant physiology that might determine the potential risk. Hence, we determine which parameters require special attention for more accurate pollution impact assessment and more generally for describing gas exchange response to a changing climate that will influence biogeochemical cycling. This might also be one key to solve long-standing issues of earth system models (ESM) which often lack PFTs specifically representing subarctic conditions in their land surface models \citep{GMD:Poulter2015,JAMES:Lawrence2019}.

The observation site Svanhovd and available observational data are presented in Section~\ref{sec:data}. From these data, we derive multi-annual averages of essential climate variables (temperature, precipitation, global irradiance, and ozone) and evaluate the deviation of 2018/19 from this norm (Section~\ref{sec:stats}). In Section~\ref{sec:do3se}, we use the $\mathrm{DO_3SE}$ model to estimate the ozone uptake by natural and semi-natural vegetation to investigate the main environmental drivers for projected ozone damage risk in 2018. To this end, we derive bespoke parameterizations for subarctic species to assess over- and underestimations arising from the use of generic parameterizations established from less climatically extreme conditions. We will summarize our results in Section~\ref{sec:conc} and discuss how this research can support improvements in future assessments of pollution risk and more generally biogeochemical cycling.

@@

We performed ozone concentration measurements in the subarctic, during the growing seasons of 2018 and 2019. Our chosen location was Svanhovd ($69.45\,\unit{^\circ N}$, $30.03\,\unit{^\circ E}$, $30\,\unit{m\,a.s.l.}$) for several reasons. There were ozone measurements there in the past, there is an agrometeorological station measuring many of the climate variables we needed, e.g. $2\,\unit{m}$ temperature, precipitation, global irradiance and soil temperature at different depths. We collected atmospheric monitoring and ozone monitoring data taken at Svanhovd over the past $35\,\unit{years}$. There is also ongoing national surveillance of air pollution (\chem{NO_x}, \chem{SO_x}, PM, radiological species). Most of these measurements are performed by the Norwegian Institute for Air Research (NILU) and they also performed our ozone measurements.\\

All measured essential climate variables for the 2018/19 growing season were obtained through \href{luftkvalitet.no}{luftkvalitet.no} operated by \citet{NILU_AIRQ} (last accessed May 2020). Our ozone monitoring data have been added to luftkvalitet.no and the EBAS database operated by \citet{NILU_EBAS} and are thus openly accessible. Long-term ozone observation data (1986--1996) were obtained from EBAS. Note that data from before the 1990s do not follow the high-quality standard procedures implemented nowadays and have to be treated with care \citep{NILU2003}. Especially, ozone monitors did not undergo regular re-calibration. This probably led to drifts in the observed data and may impose a false trend which shall not be of concern for us as we are interested in the seasonality.
The long-term data sets of agrometeorological variables including temperature and precipitation are available from September 1992 to the present day \citep[LandbruksMeteorologiske Tjeneste][note the station name here is Pasvik]{LMT_NIBIO}.

To relate local ozone concentration \chem{[O_3]} with visible damage on ozone sensitive plants, both an ozone bio-monitor (referred to as ozone garden) and a conventional ozone monitor using the existing atmospheric monitoring infrastructure had been installed during the 2018/19 growing season at Svanhovd. We report \chem{[O_3]} as volume mixing ratios (VMR) in \unit{ppb} for easier comparison with exposure-based risk assessment. The locations of the atmospheric monitoring site and the ozone garden are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a. An ozone garden consists of selected plant species that are sensitive to ozone and are likely to display visible injuries. Cultivated species were, e.g. clovers (\emph{Trifolium repens}, \emph{Trifolium pratense}), tobacco (\emph{Nicotiana tabacum}, cultivars Bel-W3, Bel-B and Bel-C), and potato (\emph{Solanum tuberosum}). As shown in Fig.~\ref{fig:svanhovd_research_station}b, the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we qualitatively observed visible ozone damage on semi-natural vegetation (clover) and crops (tobacco). In contrast, no such visible damage was found on the clovers in 2019, although it was observed on the sensitive tobacco cultivar (Bel-W3). This indicates that the vegetation was more affected by ozone in 2018 than in 2019.

@@

\section{Statistical analysis}
\label{sec:stats}
In this section, we assess the climate conditions at Svanhovd. To this end, we compute a multi-annual average (referred to as climatology) for each environmental key variable (temperature, precipitation, global irradiance) and ground-level ozone. We evaluate the statistical significance of divergences from the norm in these variables (referred to as anomalies) in 2018/19. Concerning ozone, we performed a bias correction accounting for an increase in tropospheric background \chem{[O_3]} and cross-calibrated the climatology with long-term observations at other sites in northern Fennoscandia as described in \citet{ACPD:Falk2021}.

@@

Svanhovd is located within the subarctic climate zone. The 1992--2012 climatologies of temperature and precipitation are shown in a combined climate diagram (Fig.~\ref{fig:climatediagram}). Monthly averaged $2\,\unit{m}$ temperatures (red line) are displayed with a standard error of the mean (SE) as error band. We chose SE in this case due to the large interannual variability in temperature reflected in the standard deviation. Averaged monthly accumulated precipitation (blue bars) is shown with standard deviation as error bars. Months with average temperatures below freezing are denoted with a star. As can be seen, temperatures stay below freezing for 5 consecutive months, while only 2 months breach $10\,\unit{^\circ C}$ regularly (July, August), satisfying the conditions for K\"{o}ppen's climate classification of a regular subarctic climate (Dfc) \citep[e.g.][]{SD:Beck2018}. The highest monthly average temperature is $(13.1\pm 1.1)\,\unit{^\circ C}$ in July and the lowest $(-12.8\pm 2.0)\,\unit{^\circ C}$ in February. The coldest measured temperature was $-45.2\,\unit{^\circ C}$ in January 27, 1999, while the highest temperature ($29.4\,\unit{^\circ C}$) occurred in July 16 the same year.

@@

The average accumulated monthly precipitation ($\sum_d\mathrm{Precip}$) indicates that winter and spring (November--April, except for March) are relatively dry ($\sum_d\mathrm{Precip} < 20\,\unit{mm}$). March precipitation is primarily snow and will influence the start of the growing season. The driest month is January with $\sum_d\mathrm{Precip} = (16.7\pm 3.0)\,\unit{mm}$. The most precipitation occurs in the summer months, with a $\sum_d\mathrm{Precip} = (58.5\pm 9.2)\,\unit{mm}$ in August. The average annual accumulated precipitation ($\sum_m\mathrm{Precip}$) given with standard error of mean is $(383\pm 86)\,\unit{mm}$. Precipitation shows a considerable interannual variability.

The climatology for global irradiance ($Q_0$, not shown) approaches zero in December/January (polar night) and reaches its maximum in June/July (midnight sun conditions). We select the highest $Q_0$ for each day (noon) and average each over the whole observation period. Thus, we find the highest averaged daily maximum ($\left<Q_0^\mathrm{max}\right>$) which amounts to $800\,\unit{W\,m^{-2}}$. Similarly, we diurnal mean $Q_0$ (including night) and find the maximum of the averaged daily mean ($\left<Q_0^\mathrm{mean}\right>$) $450\,\unit{W\,m^{-2}}$ (regularly reached in May--July). By selecting the lowest daily $Q_0$ (including night), we compute the maximum of the averaged daily minimum ($\left<Q_0^\mathrm{min}\right>$) about $200\,\unit{W\,m^{-2}}$. This indicates that a considerable amount of light is also available during night.

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{ozone_climatology_svanvik_obs_spline}
  %\includegraphics[width=8.3cm]{fig05}
  \caption{Svanhovd ozone climatology. The density distribution is shown together with a multi-annual average of daily \chem{[O_3]} (climatology). A spline has been fitted through daily mean \chem{[O_3]}. The climatology of daily mean ozone is shown together with $1\,\sigma$ standard deviation and standard error of the mean.}
  \label{fig:ozone_climatology_fenoscandic_obs_spline}
\end{figure}

@@

We deduced the dominant species on-site at Svanhovd and from Fig.~\ref{fig:svanhovd_research_station} and found perennial grassland, birch (generalized as deciduous trees), and Scots pine (generalized as coniferous trees). Generic parameters for these PFTs are derived from \citet{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017} and will be referred to as mapping manual (MM) parameterizations in the following. Our parameterization of coniferous trees is based on the MM's Boreal Norway spruce, deciduous trees on the MM's silver birch, and perennial grassland on MM's perennial grassland for central Europe. For a comprehensive list of generic MM model parameters, consult Supplement~Table~S1.

@@

\begin{acknowledgements}
  We thank the NIBIO Environment Centre Svanhovd for their support.
  Bj{\o}rg Rognerud for processing SeNorge.no data with respect for begin of growing season.
  Tore Flatlandsmo Berglen for hourly pressure data from Svanhovd.
  This work was supported by the Research Council of Norway (NRC) grand No. 268073.
\end{acknowledgements}
